---
title: "Arguments"
author: "YuKi"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Arguments}
  %\VignetteEngine{knitr::rmarkdown}
---
# Arguments
## Paralell
```{r eval=FALSE}
binaryRL::rcv_d(
  ...,
  funcs = c("my_util_func", "my_rate_func", "my_expl_func", "my_prob_func"),
  nc = 4,
  ...
)

binaryRL::fit_p(
  ...,
  funcs = c("my_util_func", "my_rate_func", "my_expl_func", "my_prob_func"),
  nc = 4,
  ...
)
```
Both `rcv_d` and `fit_p` support parallel computation, meaning they can fit multiple participants' datasets simultaneously as long as `nc > 1`.  
Since parallel execution runs in a separate environment, if you have customized any of the four core functions, you must explicitly pass the function names to `binaryRL` via the `funcs` argument.

<!---------------------------------------------------------->

## Utility Function
The subjective value of objective rewards is a topic that requires discussion, as different scholars may have different perspectives. This can be traced back to the `Stevens's Power Law`. In this model, you can customize your utility function. By default, I use a power function based on Stevens' power law to model the relationship between subjective and objective value.

According to **Kahneman's Prospect Theory**, individuals exhibit distinct utility functions for **gains** and **losses**. Referencing Nilsson et al. (2012), we have implemented the model below. By replacing `util_func` with the specified form that follows, you can enable the model to run a utility function based on **Kahneman's Prospect Theory**.


```{r eval=FALSE}
func_gamma <- function(
  value, utility, reward, occurrence, var1, var2, gamma, alpha, beta
){
  # Stevens's Power Law
  if (length(gamma) == 1) {
    gamma <- as.numeric(gamma)
    utility <- sign(reward) * (abs(reward) ^ gamma)
  }
  # Prospect Theory
  else if (length(gamma) == 2 & reward < 0) {
    gamma <- as.numeric(gamma[1])
    beta <- as.numeric(beta)
    
    utility <- beta * sign(reward) * (abs(reward) ^ gamma)
  }
  else if (length(gamma) == 2 & reward >= 0) {
    gamma <- as.numeric(gamma[2])
    beta <- 1
    
    utility <- beta * sign(reward) * (abs(reward) ^ gamma)
  }
  else {
    utility <- "ERROR" 
  }
  return(list(gamma, utility))
}
```

**Reference**  
Kahneman, D., & Tversky, A. (2013). Prospect theory: An analysis of decision under risk. In Handbook of the fundamentals of financial decision making: Part I (pp. 99-127).  https://doi.org/10.1142/9789814417358_0006  
Nilsson, H., Rieskamp, J., & Wagenmakers, E. J. (2011). Hierarchical Bayesian parameter estimation for cumulative prospect theory. *Journal of Mathematical Psychology, 55*(1), 84-93. https://doi.org/10.1016/j.jmp.2010.08.006

<!---------------------------------------------------------->

## Initial Value and Learning Rate
In `run_m`, there is an argument called `initial_value`. Considering that the initial value has a significant impact on the parameter estimation of the **learning rates ($\eta$)** When the initial value is not set (`initial_value = NA`), it is taken to be the reward received for that stimulus the first time.

> "Comparisons between the two learning rates generally revealed a positivity bias ($\alpha_{+} > \alpha_{-}$)"  
> "However, that on some occasions, studies failed to find a positivity bias or even reported a negativity bias ($\alpha_{+} < \alpha_{-}$)."  
> "Because Q-values initialization markedly affect learning rate and learning bias estimates."

```{r eval=FALSE}
binaryRL::run_m(
  ...,
  initial_value = NA,
  ...
)
```

**Reference**  
Palminteri, S., & Lebreton, M. (2022). The computational roots of positivity and confirmation biases in reinforcement learning. *Trends in Cognitive Sciences, 26*(7), 607-621. https://doi.org/10.1016/j.tics.2022.04.005

## Exploration–Exploitation Trade-off  
### Initial Value
> "Initial action values can also be used as a simple way of encouraging exploration. Suppose that instead of setting the initial action values to zero, as we did in the 10-armed testbed, we set them all to +5. Recall that the q(a) in this problem are selected from a normal distribution with mean 0 and variance 1. An initial estimate of +5 is thus wildly optimistic. But this optimism encourages action-value methods to explore."

**Reference**  
Sutton, R. S., & Barto, A. G. (2014, 2015). Reinforcement Learning: An Introduction (2nd ed). Cambridge: MIT press.  

### Exploration Strategy
Participants in the experiment may not always choose based on the value of the options, but instead select randomly on some trials. This is known as exploration strategy. You can implement an $\epsilon$-first model by setting the `threshold` parameter in the `run_m`. For instance, if the threshold is set to `threshold = 20` (The default value is set to 1), it means that participants will choose completely randomly until trial number 20, after which their choices will be based on value. 

```{r eval=FALSE}
# epsilon-first
run_m(
  ...
  threshold = 20,
  epsilon = NA,
  lambda = NA
  ...
)
```

The $\epsilon$-greedy strategy is commonly employed in reinforcement learning models. With this approach (`epsilon = 0.1`), the participant has a 10% probability of randomly selecting an option and a 90% probability of choosing based on the currently learned value of the options.

```{r eval=FALSE}
# epsilon-greedy
run_m(
  ...
  threshold = 1,
  epsilon = 0.1,
  lambda = NA
  ...
)
```

You can also create an $\epsilon$-decreasing exploration strategy by setting `lambda` instead of `epsilon`. In this model, the probability of participants choosing randomly will decrease as the trial number increases.

```{r eval=FALSE}
# epsilon-decreasing
run_m(
  ...
  threshold = 1,
  epsilon = NA,
  lambda = 0.1
  ...
)
```

**Reference**  
Namiki, N., Oyo, K., & Takahashi, T. (2014, December). How do humans handle the dilemma of exploration and exploitation in sequential decision making?. In Proceedings of the 8th International Conference on Bioinspired Information and Communications Technologies (pp. 113-117). https://doi.org/10.4108/icst.bict.2014.258045  
<!---------------------------------------------------------->

### Upper-Confidence-Bound
The $\pi$ parameter in Upper-Confidence-Bound (UCB) Action Selection, which is c in the book, is a parameter used to control the ratio of the bias value given to less-selected options. A larger value of this parameter will assign a greater bias to options that have been chosen infrequently. This value is set to 0.001 by default, ensuring that the model does not completely ignore any option. Once an option has been selected once, the adjustment to its value bias approaches almost zero. If the rewards in your experiment are very small or very large, you should adjust this value accordingly.

```{r eval=FALSE}
run_m(
  ...
  pi = 0.1,
  ...
)
```

**Reference**  
Sutton, R. S., & Barto, A. G. (2014, 2015). Reinforcement Learning: An Introduction (2nd ed). Cambridge: MIT press.  
Sacu, S., Dubois, M., Hezemans, F. H., Aggensteiner, P. M., Monninger, M., Brandeis, D., ... & Holz, N. E. (2024). Early-life adversities are associated with lower expected value signaling in the adult brain. Biological Psychiatry, 96(12), 948-958. https://doi.org/10.1016/j.biopsych.2024.04.005  

### Soft-Max Function
During the recovery process, the last element of both `simulate_lower` and `simulate_upper` corresponds to the $\tau$ parameter used in the softmax function.

- `simulate_lower` represents a fixed positive increment applied to all $\tau$ values.   
if this value is set to 1, it means that 1 is added to every $\tau$ during simulation.

- `simulate_upper` specifies the rate parameter of an exponential distribution from which $\tau$ is sampled.  
if this value is 1, then $\tau$ is drawn from an exponential distribution with a rate of 10, i.e., $\tau \sim \text{Exp}(1)$.


```{r eval=FALSE}
binaryRL::rcv_d(
  ...
  simulate_lower = list(c(0, 1), c(0, 0, 1), c(0, 0, 1)),
  simulate_upper = list(c(1, 1), c(1, 1, 1), c(1, 1, 1)),
  ...
)
```

**Reference**  
Wilson, R. C., & Collins, A. G. (2019). Ten simple rules for the computational modeling of behavioral data. *Elife*, 8, e49547. https://doi.org/10.7554/eLife.49547

<!---------------------------------------------------------->

## Model Fit

LL represents the similarity between the model's choices and human choices. The larger this value, the better the model.

$$
LL = \sum B_{L} \times \log P_{L} + \sum B_{R} \times \log P_{R}
$$   

$$
AIC =  - 2 LL + 2 k
$$

$$
BIC =  - 2 LL + k \times \log n
$$ 

*NOTE:* $B_{L}$ and $B_{R}$ the option that the subject chooses. ($B_{L} = 1$: subject chooses the left option; $B_{R} = 1$: subject chooses the right option); $P_{L}$ and $P_{R}$ represent the probabilities of selecting the left or right option, as predicted by the reinforcement learning model. ${k}$ the number of free parameters in the model; ${n}$ represents the total number of trials in the paradigm.

**Reference**  
Hampton, A. N., Bossaerts, P., & O'doherty, J. P. (2006). The role of the ventromedial prefrontal cortex in abstract state-based inference during decision making in humans. *Journal of Neuroscience, 26*(32), 8360-8367. https://doi.org/10.1523/JNEUROSCI.1010-06.2006

