---
title: "Markov Chain Monte Carlo"
author: "YuKi"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MCMC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The core functions of `binaryRL` have already been rewritten in `Rcpp`. This optimization reduces the runtime to roughly one-tenth of the previous R-based for-loop implementation. As a result, running MCMC with `LaplacesDemon` on `binaryRL` is very fast, while still allowing users to easily define and customize their reinforcement learning models.

## Install Package
```{r eval=FALSE}
install.packages("LaplacesDemon")
```

## Simulating Data

```{r eval=FALSE}
list_simulated <- binaryRL::simulate_list(
  data = binaryRL::Mason_2024_G2,
  id = 1,
  n_params = 2,
  n_trials = 360,
  
  obj_func = binaryRL::TD,
  rfun = list(
    eta = function() { stats::rbeta(n = 1, shape1 = 2, shape2 = 2) },
    tau = function() { stats::rexp(n = 1, rate = 1) }
  ),
  iteration = 1100 # = [1000(drop) + 100(test)]]
)

list_drop <- list_simulated[1:1000]
list_test <- list_simulated[1001:1100]

for (i in seq_along(list_test)) {
  list_test[[i]][["data"]]$Subject <- i
}

test_data <- dplyr::bind_rows(purrr::map(list_test, "data"))
```

## Step 1: Object Function

### Argument 1: parm

Similar to other frameworks designed for individual-level analysis, the primary argument to the `LaplacesDemon` model function is a single vector of free parameters. Crucially, this vector does not represent the parameters for a single "black-box" function call. Instead, it is a concatenated vector containing the complete set of free parameters for every subject being modeled. Consequently, its total length is the number of free parameters per subject multiplied by the number of subjects.

### Argument 2: Data

A key convenience of `LaplacesDemon` is that the user-defined Model function can accept a `Data` argument. This feature provides a clean and explicit mechanism for passing supplementary information—such as prior distributions, trial counts, or policy settings—directly to the model. Consequently, users can avoid more cumbersome methods, like environment binding, and instead bundle all necessary data and constants into a single, well-organized list.

```{r eval=FALSE}
Data <- list(
  # The prepared dataframe containing the experimental data
  df = test_data,
  # A vector of unique subject IDs to loop over during modeling
  id = unique(test_data$Subject),
  # Need to constrain the free parameters with Laplace
  priors = NULL,
  # The number of free parameters per subject (eta and tau)
  n_params = 2,
  # The total number of trials per subject in the experiment
  n_trials = 360,
  # The operational mode for the binaryRL::run_m function, set to "fit"
  mode = "fit",
  # The learning policy for binaryRL::run_m, set to "off-policy"
  policy = "off",
  # A vector of descriptive names for each parameter across all subjects
  parm.names = LaplacesDemon::as.parm.names(
    list(
      outer(
        X = c("eta", "tau"),
        Y = 1:length(unique(test_data$Subject)),
        FUN = paste, sep = ""
      )
    )
  ),
  # The names of variables to monitor during the MCMC run (Log Posterior)
  mon.names = "LP",
  # The total number of subjects
  N = length(unique(test_data$Subject))
)

# Add Block [eta, tau] 
Data$B <- lapply(1:length(unique(test_data$Subject)), function(i) {
  c((i - 1) * Data$n_params + 1, (i - 1) * Data$n_params + 2)
})
```

## Step 2: Defining Model

Create a model function that accepts `parm` and `Data` as its arguments, with a return value that conforms to the structure required by `LaplacesDemon.` Within this function, you need to calculate the sum of the log-likelihoods and the sum of the log-posteriors across all subjects by using specified prior distributions. 

```{r eval=FALSE}
Model <- function(parm, Data) {
  
  # Initialize accumulators for sum of log-likelihood and log-posterior
  sum_LL <- 0
  sum_LP <- 0
  
  # Loop through each subject ID provided in the Data list
  results <- parallel::parLapply(
    cl = cl,
    X = as.vector(Data$id),
    fun = function(i) {
      
      # Read unconstrained params for the current subject
      eta_unconstrained <- parm[(i - 1) * 2 + 1]
      tau_unconstrained <- parm[(i - 1) * 2 + 2]
      
      # Convert to constrained params
      eta <- stats::plogis(eta_unconstrained)
      tau <- exp(tau_unconstrained)
      
      # Run the model for the current subject
      res <- binaryRL::run_m(
        data = Data$df,
        id = i, 
        eta = eta,
        tau = tau,
        priors = Data$priors,
        n_params = Data$n_params,
        n_trials = Data$n_trials,
        mode = Data$mode,
        policy = Data$policy
      )
      
      # Calculate the log-posterior components for this subject
      LP_eta <- stats::dbeta(x = eta, shape1 = 2, shape2 = 2, log = TRUE)
      LP_tau <- stats::dexp(x = tau, rate = 1, log = TRUE)
      
      # Return a list with the results for this subject
      return(c(ll = res$ll, lp = LP_eta + LP_tau))
    }
  )
  
  # After the loop, aggregate the results from the list
  sums <- rowSums(do.call(cbind, results))
  # Extract all log-likelihoods and sum them up
  sum_LL <- sums["ll"]
  
  # Extract all subject-wise log-posteriors and sum them up
  sum_LP <- sums["lp"]
  
  LP <- sum_LL + sum_LP
  
  # Return a list in the format required by LaplacesDemon
  return(
    list(
      LP = LP,
      Dev = -2 * sum_LL,
      Monitor = c(LP = LP),
      yhat = NULL,
      parm = parm
    )
  )
}
```

## Step 3: Test Model

```{r eval=FALSE}
set.seed(123)
cl <- parallel::makeCluster(10) 
parallel::clusterEvalQ(cl = cl, expr = library(binaryRL))
Initial.Values <- stats::runif(length(Data$id) * Data$n_params)
parallel::clusterExport(cl, varlist = c("Initial.Values"))

profvis::profvis({Model(
  parm = Initial.Values, 
  Data = Data
)})
Model(
  parm = Initial.Values, 
  Data = Data
)
```

## Step 4: Run LaplacesDemon

```{r eval=FALSE}
Fit <- LaplacesDemon::LaplacesDemon(
  Model = Model,
  Data = Data,
  Packages = "binaryRL",
  # Set initial values (# of free params * # of subjects)
  Initial.Values = Initial.Values,
  # The relationship between `Algorithm` and `Specs` is special
  # Refer to the LaplacesDemon help documents for details
  Algorithm = "MWG", Specs = list(B = Data$B),
  #Algorithm = "NUTS", Specs = list(A = 500, delta = 0.6, epsilon = NULL, Lmax = Inf),
  # The total number of iterations to perform in each chain.
  Iterations = 10,
  # The thinning interval. A value of n keeps one sample every n-th iteration.
  Thinning = 1,
  # Logs for every ? iterations
  Status = 1
)

parallel::stopCluster(cl)

LaplacesDemon::Consort(Fit)
```

```
#############################################################
# Consort with Laplace's Demon                              #
#############################################################
Call:
LaplacesDemon::LaplacesDemon(
  Model = Model, Data = Data, Initial.Values = Initial.Values,
  Iterations = 10, Status = 1, Thinning = 1, 
  Algorithm = "MWG", Specs = list(B = Data$B), Packages = "binaryRL"
)

Acceptance Rate: 0.79
Algorithm: Metropolis-within-Gibbs
Covariance Matrix: (NOT SHOWN HERE; diagonal shown instead)
     [1,1]      [2,1]      [1,2]      [2,2]      [1,3]      [2,3]
0.02835066 0.02835066 0.02835066 0.02835066 0.02835066 0.02835066 
...

Covariance (Diagonal) History: (NOT SHOWN HERE)
Deviance Information Criterion (DIC):
           All Stationary
Dbar  52925.39   52925.39
pD   156998.62  156998.62
DIC  209924.01  209924.01
Initial Values:
  [1] 0.2875775201 0.7883051354 0.4089769218 0.8830174040 0.9404672843 
...

Iterations: 11
Log(Marginal Likelihood): NA
Minutes of run-time: 34.03
Model: (NOT SHOWN HERE)
Monitor: (NOT SHOWN HERE)
Parameters (Number of): 200
Posterior1: (NOT SHOWN HERE)
Posterior2: (NOT SHOWN HERE)
Recommended Burn-In of Thinned Samples: 0
Recommended Burn-In of Un-thinned Samples: 0
Recommended Thinning: 10
Specs: (NOT SHOWN HERE)
Status is displayed every 1 iterations
Summary1: (SHOWN BELOW)
Summary2: (SHOWN BELOW)
Thinned Samples: 11
Thinning: 1


Summary of All Samples
                  Mean           SD         MCSE       ESS            LB        Median            UB
[1,1]     1.767571e-01 5.826327e-02 4.213680e-02  1.435352  9.628475e-02  1.892334e-01  2.631710e-01
[2,1]     7.222023e-01 4.039864e-02 2.679293e-02  2.264387  6.770064e-01  7.335398e-01  7.885450e-01
[1,2]     4.306596e-01 2.253195e-02 1.059517e-02 11.000000  4.032490e-01  4.310931e-01  4.655382e-01
[2,2]     9.438012e-01 3.990277e-02 2.470063e-02  2.168657  8.731643e-01  9.635338e-01  9.816604e-01
[1,3]     8.858296e-01 3.415728e-02 1.745856e-02  3.874139  8.214295e-01  9.014520e-01  9.143744e-01
[2,3]    -2.304814e-02 4.930958e-02 3.747143e-02  1.308741 -8.159860e-02 -3.981861e-02  4.360556e-02
[1,4]     5.211726e-01 3.274488e-02 1.998962e-02  2.233907  4.862525e-01  5.261202e-01  5.694293e-01
[2,4]     8.254009e-01 2.712795e-02 1.599573e-02  2.983164  7.853919e-01  8.421988e-01  8.528547e-01
[1,5]     5.033317e-01 4.083967e-02 3.047860e-02  1.493804  4.436491e-01  5.266891e-01  5.513182e-01
[2,5]     4.498422e-01 2.069964e-02 1.162955e-02 11.000000  4.110313e-01  4.566147e-01  4.718698e-01
...
 [ reached getOption("max.print") -- omitted 60 rows ]


Summary of Stationary Samples
                  Mean           SD         MCSE       ESS            LB        Median            UB
[1,1]     1.767571e-01 5.826327e-02 4.213680e-02  1.435352  9.628475e-02  1.892334e-01  2.631710e-01
[2,1]     7.222023e-01 4.039864e-02 2.679293e-02  2.264387  6.770064e-01  7.335398e-01  7.885450e-01
[1,2]     4.306596e-01 2.253195e-02 1.059517e-02 11.000000  4.032490e-01  4.310931e-01  4.655382e-01
[2,2]     9.438012e-01 3.990277e-02 2.470063e-02  2.168657  8.731643e-01  9.635338e-01  9.816604e-01
[1,3]     8.858296e-01 3.415728e-02 1.745856e-02  3.874139  8.214295e-01  9.014520e-01  9.143744e-01
[2,3]    -2.304814e-02 4.930958e-02 3.747143e-02  1.308741 -8.159860e-02 -3.981861e-02  4.360556e-02
[1,4]     5.211726e-01 3.274488e-02 1.998962e-02  2.233907  4.862525e-01  5.261202e-01  5.694293e-01
[2,4]     8.254009e-01 2.712795e-02 1.599573e-02  2.983164  7.853919e-01  8.421988e-01  8.528547e-01
[1,5]     5.033317e-01 4.083967e-02 3.047860e-02  1.493804  4.436491e-01  5.266891e-01  5.513182e-01
[2,5]     4.498422e-01 2.069964e-02 1.162955e-02 11.000000  4.110313e-01  4.566147e-01  4.718698e-01
...
 [ reached getOption("max.print") -- omitted 60 rows ]

Demonic Suggestion

Due to the combination of the following conditions,

1. Metropolis-within-Gibbs
2. The acceptance rate (0.79) is above 0.5.
3. At least one target MCSE is >= 6.27% of its marginal posterior
   standard deviation.
4. At least one target distribution has an effective sample size
   (ESS) less than 100. The worst mixing chain is: [2,73] (ESS=1.121322).
5. Each target distribution became stationary by
   1 iteration.

Quantiles of Absolute Posterior1 Correlation:
          0%          25%          50%          75%         100% 
8.216533e-06 3.237191e-01 5.822562e-01 7.866556e-01 1.000000e+00 

Possibly excessive posterior correlation for a componentwise algorithm.

Laplace's Demon has not been appeased, and suggests
copy/pasting the following R code into the R console,
and running it.

Initial.Values <- as.initial.values(Fit)
Fit <- LaplacesDemon(Model, Data=Data, Initial.Values,
     Covar=Fit$Covar, Iterations=110, Status=0, Thinning=10,
     Algorithm="AMWG", Specs=list(B=Fit$Specs$B, n=11, Periodicity=13))

Laplace's Demon is finished consorting.
```
