---
title: "Markov Chain Monte Carlo"
author: "YuKi"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MCMC}
  %\VignetteEngine{knitr::rmarkdown}
---
`LaplacesDemon` provides a user-friendly interface for performing MCMC sampling on arbitrary R functions, which are often treated as "black boxes." This makes it exceptionally convenient for conducting Bayesian inference on existing, complex models without significant refactoring.

This approach stands in contrast to probabilistic programming languages like `Stan`. `Stan` delivers superior computational performance by first requiring the user to rewrite the entire model logic in its own high-level statistical language. The `Stan` compiler then translates this model into highly optimized `C++` code, which eliminates the overhead of an interpreted language like R.

The primary drawback and performance bottleneck of using `LaplacesDemon` with an R-based model function lies here: while the sampler itself is efficient, it must repeatedly call the R function at every MCMC iteration. As observed with the binaryRL package, this process becomes extremely inefficient for complex models, as the runtime is dominated by the interpreted R code within the "black-box" function.

To address this critical performance issue, the future direction for `binaryRL` will be to rewrite the core `run_m` function in `C++` (likely using the `Rcpp` package). By compiling the model's core logic, we can drastically reduce the evaluation time of the objective function, thus making large-scale Bayesian inference with samplers like `LaplacesDemon` both practical and efficient.

## Example Data

```{r}
test_data <- binaryRL::Mason_2024_Exp2
# Ensure the Subject ID is a numeric value for easier indexing
test_data$Subject <- as.numeric(test_data$Subject)
# Filter the dataset to include only subjects 1 through 5
test_data <- test_data[test_data$Subject <= 5, ]

unique(test_data$Subject)
```

## Object Function

### Argument 1: parm

Similar to other frameworks designed for individual-level analysis, the primary argument to the `LaplacesDemon` model function is a single vector of free parameters. Crucially, this vector does not represent the parameters for a single "black-box" function call. Instead, it is a concatenated vector containing the complete set of free parameters for every subject being modeled. Consequently, its total length is the number of free parameters per subject multiplied by the number of subjects.

### Argument 2: Data

A key convenience of `LaplacesDemon` is that the user-defined Model function can accept a `Data` argument. This feature provides a clean and explicit mechanism for passing supplementary information—such as prior distributions, trial counts, or policy settings—directly to the model. Consequently, users can avoid more cumbersome methods, like environment binding, and instead bundle all necessary data and constants into a single, well-organized list.

```{r}
Data <- list(
  # The prepared dataframe containing the experimental data
  df = test_data,
  # A vector of unique subject IDs to loop over during modeling
  id = unique(test_data$Subject),
  # A list of prior distributions for the parameters
  priors = list(
    # A uniform prior for eta, constraining it between 0 and 1
    eta = function(x) { stats::dunif(x, min = 0, max = 1, log = TRUE) },
    # An exponential prior for tau, with a rate of 1
    tau = function(x) { stats::dexp(x, rate = 1, log = TRUE) }
  ),
  # The number of free parameters per subject (eta and tau)
  n_params = 2,
  # The total number of trials per subject in the experiment
  n_trials = 360,
  # The operational mode for the binaryRL::run_m function, set to "fit"
  mode = "fit",
  # The learning policy for binaryRL::run_m, set to "off-policy"
  policy = "off",
  # A vector of descriptive names for each parameter across all subjects
  parm.names = LaplacesDemon::as.parm.names(
    list(
      # Creates names like "[1,1]", "[1,2]", "[2,1]", "[2,2]", etc.
      outer(
        X = c("eta", "tau"),
        Y = 1:length(unique(test_data$Subject)),
        FUN = paste, sep = ""
      )
    )
  ),
  # The names of variables to monitor during the MCMC run (Log Posterior)
  mon.names = "LP",
  # The total number of subjects
  N = length(unique(test_data$Subject))
)
```

### Defining Object Function

Create a model function that accepts `parm` and `Data` as its arguments, with a return value that conforms to the structure required by `LaplacesDemon.` Within this function, you need to calculate the sum of the log-likelihoods and the sum of the log-posteriors across all subjects by using specified prior distributions. These priors effectively define the valid domain for the parameters. If `LaplacesDemon` attempts to use a parameter value that falls outside the support of its prior distribution, the function will return negative infinity.

```{r}
Model <- function(parm, Data) {
  
  # Initialize accumulators for sum of log-likelihood and log-posterior
  sum_LL <- 0
  sum_LP <- 0
  
  # Loop through each subject ID provided in the Data list
  for (i in as.vector(Data$id)) {
    
    eta <- parm[(i - 1) * 2 + 1]
    tau <- parm[(i - 1) * 2 + 2]
    
    res <- binaryRL::run_m(
      data = Data$df,
      id = Data$id[i],
      eta = eta,
      tau = tau,
      priors = Data$priors,
      n_params = Data$n_params,
      n_trials = Data$n_trials,
      mode = Data$mode,
      policy = Data$policy
    )
    
    # Add the log-likelihood of the current subject to the total
    sum_LL <- sum_LL + res$ll
    # Add the log-posterior of the current subject to the total
    sum_LP <- sum_LP + res$lpo
  }
  
  # Return a list in the format required by LaplacesDemon
  return(
    list(
      LP = sum_LP,
      Dev = -2 * sum_LL,
      Monitor = c(LP = sum_LP),
      yhat = NULL,
      parm = parm
    )
  )
}
```

### Test Object Function

```{r}
Initial.Values <- stats::runif(Data$N * Data$n_params)
Model(parm = Initial.Values, Data = Data)
```

## Run LaplacesDemon

```{r eval=FALSE}
Fit <- LaplacesDemon::LaplacesDemon.hpc(
  Model = Model,
  Data = Data,
  Packages = "binaryRL",
  # Set initial values (# of free params * # of subjects)
  Initial.Values = stats::runif(length(Data$id) * 2),
  # The relationship between `Algorithm` and `Specs` is special
  # Refer to the LaplacesDemon help documents for details
  Algorithm = "MWG", Specs = list(B = NULL),
  # The total number of iterations to perform in each chain.
  Iterations = 10,
  # The thinning interval. A value of n keeps one sample every n-th iteration.
  Thinning = 1,
  # The number of parallel chains to run (Chains == CPUs).
  Chains = 16, 
  # The number of CPU cores to use (CPUs == Chains).
  CPUs = 16
)

print(Fit, Burnin = 1)
```

```
[[1]]
Call:
LaplacesDemon(Model = Model, Data = Data, Initial.Values = Initial.Values[x, 
    ], Covar = Covar, Iterations = Iterations, Status = Status, 
    Thinning = Thinning, Algorithm = Algorithm, Specs = Specs, 
    Debug = Debug, LogFile = paste(LogFile, ".", x, sep = ""))

Acceptance Rate: 0.06364
Algorithm: Metropolis-within-Gibbs
Covariance Matrix: (NOT SHOWN HERE; diagonal shown instead)
    [1,1]     [2,1]     [1,2]     [2,2]     [1,3]     [2,3]     [1,4]     [2,4]     [1,5]     [2,5] 
0.5670132 0.5670132 0.5670132 0.5670132 0.5670132 0.5670132 0.5670132 0.5670132 0.5670132 0.5670132 

Covariance (Diagonal) History: (NOT SHOWN HERE)
Deviance Information Criterion (DIC):
          All Stationary
Dbar 2275.985   2275.985
pD     74.704     74.704
DIC  2350.690   2350.690
Initial Values:
 [1] 0.09859165 0.69300856 0.75132556 0.99870695 0.34549739 0.29300554 0.66773598 0.59801099 0.70105000 0.10186876

Iterations: 11
Log(Marginal Likelihood): NA
Minutes of run-time: 3.49
Model: (NOT SHOWN HERE)
Monitor: (NOT SHOWN HERE)
Parameters (Number of): 10
Posterior1: (NOT SHOWN HERE)
Posterior2: (NOT SHOWN HERE)
Recommended Burn-In of Thinned Samples: 0
Recommended Burn-In of Un-thinned Samples: 0
Recommended Thinning: 10
Specs: (NOT SHOWN HERE)
Status is displayed every 10 iterations
Summary1: (SHOWN BELOW)
Summary2: (SHOWN BELOW)
Thinned Samples: 11
Thinning: 1


Summary of All Samples
                  Mean         SD       MCSE          ESS            LB        Median            UB
[1,1]     9.859165e-02  0.0000000 0.00000000 2.220446e-16  9.859165e-02  9.859165e-02  9.859165e-02
[2,1]     5.200845e-01  0.0000000 0.00000000 2.220446e-16  5.200845e-01  5.200845e-01  5.200845e-01
[1,2]     7.513256e-01  0.0000000 0.00000000 2.220446e-16  7.513256e-01  7.513256e-01  7.513256e-01
[2,2]     5.810195e-01  0.0000000 0.00000000 2.220446e-16  5.810195e-01  5.810195e-01  5.810195e-01
[1,3]     7.691418e-01  0.0000000 0.00000000 2.220446e-16  7.691418e-01  7.691418e-01  7.691418e-01
[2,3]     8.243759e-01  0.1762356 0.05020148 1.100000e+01  4.391324e-01  8.775129e-01  8.775129e-01
[1,4]     9.930241e-01  0.0000000 0.00000000 2.220446e-16  9.930241e-01  9.930241e-01  9.930241e-01
[2,4]     9.920084e-01  0.1306741 0.03722310 1.100000e+01  7.063603e-01  1.031408e+00  1.031408e+00
[1,5]     7.010500e-01  0.0000000 0.00000000 2.220446e-16  7.010500e-01  7.010500e-01  7.010500e-01
[2,5]     6.699087e-01  0.1883975 0.05366585 1.100000e+01  2.580797e-01  7.267127e-01  7.267127e-01
Deviance  2.275985e+03 12.2232699 3.48185156 1.100000e+01  2.272300e+03  2.272300e+03  2.302705e+03
LP       -1.141580e+03  5.6163277 1.59983535 1.100000e+01 -1.153857e+03 -1.139887e+03 -1.139887e+03


Summary of Stationary Samples
                  Mean         SD       MCSE          ESS            LB        Median            UB
[1,1]     9.859165e-02  0.0000000 0.00000000 2.220446e-16  9.859165e-02  9.859165e-02  9.859165e-02
[2,1]     5.200845e-01  0.0000000 0.00000000 2.220446e-16  5.200845e-01  5.200845e-01  5.200845e-01
[1,2]     7.513256e-01  0.0000000 0.00000000 2.220446e-16  7.513256e-01  7.513256e-01  7.513256e-01
[2,2]     5.810195e-01  0.0000000 0.00000000 2.220446e-16  5.810195e-01  5.810195e-01  5.810195e-01
[1,3]     7.691418e-01  0.0000000 0.00000000 2.220446e-16  7.691418e-01  7.691418e-01  7.691418e-01
[2,3]     8.243759e-01  0.1762356 0.05020148 1.100000e+01  4.391324e-01  8.775129e-01  8.775129e-01
[1,4]     9.930241e-01  0.0000000 0.00000000 2.220446e-16  9.930241e-01  9.930241e-01  9.930241e-01
[2,4]     9.920084e-01  0.1306741 0.03722310 1.100000e+01  7.063603e-01  1.031408e+00  1.031408e+00
[1,5]     7.010500e-01  0.0000000 0.00000000 2.220446e-16  7.010500e-01  7.010500e-01  7.010500e-01
[2,5]     6.699087e-01  0.1883975 0.05366585 1.100000e+01  2.580797e-01  7.267127e-01  7.267127e-01
Deviance  2.275985e+03 12.2232699 3.48185156 1.100000e+01  2.272300e+03  2.272300e+03  2.302705e+03
LP       -1.141580e+03  5.6163277 1.59983535 1.100000e+01 -1.153857e+03 -1.139887e+03 -1.139887e+03
```

